---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BLCOP

<!-- badges: start -->
<!-- badges: start -->
[![R-CMD-check](https://github.com/MangoTheCat/BLCOP/workflows/R-CMD-check/badge.svg)](https://github.com/MangoTheCat/BLCOP/actions)
<!-- badges: end -->

<!-- badges: end -->

The `{BLCOP}` package is an implementation of the Black-Litterman and copula opinion pooling frameworks. The Black-Litterman model was devised in 1992 by Fisher Black and Robert Litterman. Their goal was to create a systematic method of specifying and then incorporating analyst/portfolio manager views into the estimation of market parameters.

* `BLViews()` and `COPViews()` construct views objects
* `addBLViews()` and `addCOPViews()` allow more views to be added to existing objects
* `distribution()` and `mvdistribution()` create `distribution` and `mvdistribution` objects 
* `BLPosterior()` calculates the posterior distribution using the Black-Litterman model
* `COPPosterior()` calculates the posterior distribution using copula opinion pooling


## Installation

You can install the released version of BLCOP from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("BLCOP")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MangoTheCat/BLCOP")
```
## Example

```{r example, R.options = list(width = 96)}
library(BLCOP)

# For a matrix of monthly returns for 6 assets
head(monthlyReturns)

# Define a pick matrix (a vector of confidences)
pickMatrix <- matrix(c(1/2, -1, 1/2, 0, 0, 0), 
                     nrow = 1, 
                     ncol = 6)

# Create a views object
views <- BLViews(P = pickMatrix,
                 q = 0.06, 
                 confidences = 100,
                 assetNames = colnames(monthlyReturns))

# Determine the posterior distribution of these assets
BLPosterior(monthlyReturns, views, tau = 1/2, marketIndex = sp500Returns)
```
